\chapter{The Standard Model of Particle Physics}\label{sec:theory}

This chapter presents the theoretical background of this thesis through the introduction of the Standard Model of particle physics.
The Standard Model is a successful description of Nature based on the elegant foundation of group theory and abstract algebra.
These concepts are beyond the scope of this thesis, but not beyond the scope of its appendices.
A description of groups, Lie algebras, and related concepts is provided with minimal overhead in Appendix \ref{sec:math}.

The Standard Model describes many of the interactions and particles that make up the universe.
It is formulated as a \emph{quantum field theory} based on operator valued fields that operate on states belonging to a multi-dimensional Hilbert space.
It is also formulated as a \emph{gauge theory}, such that the physical predictions of the theory are invariant under certain transformations.
These transformations are representations of a Lie algebra, $\mathfrak{g}$, and are specified by continuous group parameters.
This is the algebra of the Standard Model Lie group \gsm.
The group \gsm is direct product of several subgroups.
These subgoups are listed here and a more detailed definition is provided in Appendix \ref{sec:particlePhysicsGroups}.

% Global symmetry
The first subgroup is the \poincare group $P$ of rank ten that contains translations, rotations, and boosts through spacetime.
The \poincare group itself is the semidirect product of a translation group, T(4), and the Lorentz group, SO(1;3).
The inclusion of this group means that the prediction of Standard Model are invariant under these transformations through spacetime.
The invariance under \poincare group transformations is called a \emph{global symmetry} because it is external to the fields that comprise the theory.
It forms the group of external symmetries, $\gsm^\text{external}$.

% Internal symmetry
The invariances under the remaining subgroups are called \emph{internal symmetries}, $\gsm^\text{internal}$, because they describe rotations of fields through abstract mathematical space.
There are three such subgroups in the Standard Model: U(1), SU(2), and SU(3).
The algebra of U(1) is represented by $1\times1$ complex unitary matrices.
These are invertible and have determinants equal to $\pm1$.
The algebra of SU(2) is represented by $2\times2$ complex unitary matrices with determinants equal to $+1$.
These describe rotations through the abstract mathematical space of a complex 2-sphere that preserve distance.
The algebra is spanned by the three Pauli matrices.
Finally the algebra of SU(3) is represented by $3\times3$ complex unitary matrices with determinants equal to $+1$.
The algebra is spanned by the eight Gell-Mann matrices.
The group of internal symmetries is the direct product of these three subgroups as shown in Equation \ref{eqn:smSym}.
\begin{equation}\begin{split}\label{eqn:smSym}
    \gsm^\text{internal}\equiv\text{SU}(3)\times\text{SU}(2)\times\text{U}(1)
\end{split}\end{equation} 
The theory is defined in order to be invariant under the group $\gsm=\gsm^\text{external}\times\gsm^\text{internal}$.

The dynamics of the theory are determined through a Lagrangian formulation. 
A Lagrangian density, $\mathcal{L}$, is written in terms of operator valued fields that correspond to various particles.
The dynamics of these fields are then determined by the Euler-Lagrange equation.
Further predictions of particle scattering and decay probabilities are derived from the time dependant Schrodinger equation.
An explanation as to how such measurable quantities arise from the theory is provided in Appendix \ref{sec:lagrangians}.
Since the Lagrangian density is a singlet under \gsm transformations, so are these predictions.

The subgroups of $\gsm^\text{internal}$ give rise to the dynamics of the Standard Model.
The SU(3) group is associated with the strong force and the $\text{SU}(2)\times\text{U}(1)$ is associated with the electroweak force.
These are described by individual field theories because the associated with these forces manifest at very different energy scales.
Strong interactions are described by Quantum Chromodynamics as defined in Section \ref{sec:thyQCD}.
Electroweak interactions are described by Electroweak Theory as defined in Section \ref{sec:thyEW}.

\section{Quantum Chromodynamics}\label{sec:thyQCD}

The quantum field theory of Quantum Chromodynamics (QCD) describes interactions between gluons and fermions carrying \emph{color} charge.
There are two types of elements of the theory: eight massless gluons and six quarks.
Each of these can carry color charges: quarks carry one charge, while gluons carry two.
There are three charges: red/anti-red, green/anti-green, and blue/anti-blue.

% Quarks
Quarks are described by Dirac fields with three components,
\begin{equation}\begin{split}
    q=&\begin{pmatrix}q_\text{red}&q_\text{blue}&q_\text{green}\end{pmatrix}^T.
\end{split}\end{equation}

The fields transform under SU(3) transformations,
\begin{equation}\begin{split}
    \text{SU(3):}\quad q\to&e^{i\theta^a(x)T^a}q,
\end{split}\end{equation}
where $T^a\equiv\half\lambda^a$ are eight generators based on the Gall-Mann matrices $\lambda^a$.
The Dirac field transforms under SU(3) with eight generators.
The functions $\theta^a(x)$ are eight group parameter functions of spacetime that identify the transformation.

% Gluons
The derivative of the field $q$ introduces eight terms corresponding to each gauge parameter.
In QCD, eight gluon gauge fields, $A_\mu^a$, are introduced which carry a lower spacetime index $\mu\in\{0,1,2,3\}$, and an upper representation index $a\in\{1,...,8\}$.
The fields $A_\mu^a$ transform under the adjoint representation of SU(3) as a connection:
\begin{equation}\begin{split}
    \text{SU(3):}\quad A_\mu^a\to&A_\mu^a-\frac{1}{g_s}\partial_\mu\theta^a-f^{abc}\theta^bA^c_\mu. \\
\end{split}\end{equation}
Here $f^{abc}$ are the SU(3) structure constants, $\theta^a$ are group parameters that define the transformation, and $g_s$ is a coupling constant.

The QCD Lagrangian can be written through combinations of these fields that are invariant under SU(3) transformations.
% The covariant derivative, $D_\mu\Psi_i\equiv\partial_\mu\Psi_i+igA_\mu^aT_i^{aj}\Psi_j$, is defined to allow kinetic derivatives of the fermion fields.
The covariant derivative is defined with the help of the eight gluon fields,
\begin{equation}\begin{split}
    D_\mu\equiv\partial_\mu+ig_sA_\mu^aT^a.
\end{split}\end{equation}
% checked in romao
An antisymmetric field strength tensor is defined as $F^a_{\mu\nu}\equiv\partial_\mu A^a_\nu-\partial_\nu A^a_\mu+g_sf^{abc}A_\mu^bA_\nu^c$.
The product $F^{\mu\nu,a}F^a_{\mu\nu}$ includes terms with two, three, and four $A_\mu^a$ fields.
Terms with two fields are part of the free Lagrangian, while terms with more fields describe multi-gluon vertices in the interaction Lagrangian.
The Lagrangian is given in Equation \ref{eqn:qcdLagrangian} with indices on the fermion fields $i$ and $j$ summed over the six quarks.

\begin{flalign}\label{eqn:qcdLagrangian}
    ~& &\mathcal{L}_\text{0}=                &-\frac{1}{4}(\partial_\mu A^a_\nu-\partial_\nu A^a_\mu)(\partial^\mu A^{a\nu}-\partial^\nu A^{a\mu}) & \text{free Lagrangian} \notag\\
                                                &&& +\overline{q}^i(i\slashed{\partial}-m)q_i;  \notag\\
    ~& &\mathcal{L}_{A\overline{q}q}=     &-g_sq^i\slashed{A}^aT_i^{a,j}q_j; & \text{} \notag\\
    ~& &\mathcal{L}_{AAA}=                      &-\frac{g_s}{4}f^{ab'c'}(\partial_\nu A^a_\mu-\partial_\mu A^a_\nu)A^{b'\mu}A^{c'\nu}) \notag\\
                                                &&& +\frac{g_s}{4}f^{abc}(A^b_\mu A^c_\nu)(\partial^{\nu}A^{a\mu}-\partial^\mu A^{a\nu}); \notag\\
    ~& &\mathcal{L}_{AAAA}=                     &-\frac{g_s^2}{4}(f^{abc}A^b_\mu A^c_\nu)(f^{ab'c'}A^{b'\mu}A^{c'\nu}); & \text{} \notag\\
    ~& &\mathcal{L}_\text{QCD}=&\mathcal{L}_\text{0}+\mathcal{L}_{A\overline{q}q}+\mathcal{L}_{AAAA}+\mathcal{L}_{AAA}; & \text{full Lagrangian}
\end{flalign} 
Through $\mathcal{L}_{A\overline{q}q}$, this Lagrangian includes terms that describe three-point interactions between two quarks and a gluon.
The terms $\mathcal{L}_{AAA}$ and $\mathcal{L}_{AAAA}$ describe a three- and four-point interactions involving only gluons, respectively. 
A more compact version of the QCD Lagrangian that makes use of the field strength tensor is shown in Equation \ref{eqn:qcdLagrangian2}.
\begin{equation}\begin{split}\label{eqn:qcdLagrangian2}
    \mathcal{L}_\text{QCD}=-\frac{1}{4}F^{\mu\nu a}F^a_{\mu\nu}+i\overline{q}^i\slashed{D}_\mu q_i-m\overline{q}^iq_i \\
\end{split}\end{equation}

The coupling constant $g_s$ is related to the bare coupling constant $\overline{\alpha}_s=\frac{g^2_s}{4\pi}$; however, this is not the full physical picture.
There is a particularly significant impact on the observed coupling as a result of vacuum polarization.
This is due to the gluon self-interaction terms $\mathcal{L}_{AAA}$ and $\mathcal{L}_{AAAA}$.
The value of the effective coupling $\alpha_s(Q^2)$ is a function of the momentum transfer $Q$ of an interaction, with the functional dependence given to leading order in Equation \ref{eqn:alphas}.
\begin{equation}\begin{split}\label{eqn:alphas}
    \as(Q^2)=&\frac{\as(Q_0)}{1+\as(Q_0)\beta_0\ln\frac{Q^2}{Q_0^2}} \\
\end{split}\end{equation}
Here, $\as(Q_0)$ has been measured at a particular momentum transfer $Q_0$.
The number of quark flavors accessible at the energy scale $Q$ is $N_f$, and $\beta_0=\frac{33-2N_f}{12\pi}$.
As $Q^2$ gets smaller than $Q_0^2$, the coupling $\as(Q^2)$ blows up.
This leads to the confinement of quarks and gluons in strongly bound states and explains the difficulty in observing free quarks.
Conversely, at low energy scales, the coupling constant decreases.
This leads to the phenomena of asymptotic freedom, wherein the effective strength of the strong force becomes asymptotically weak.

While QCD predicts the functional form of $\alpha_s(Q)$ with respect to the energy scale, it does not predict the absolute value.
Instead, experiments have measured $\alpha_s(Q)$ at various energy scales.
This work provides a powerful test of the theory, along with a significant input.
Measurements at energy scale of the \Z find $\alpha_s(M_{Z^0})=0.1184\pm0.0007$.
Additional measurements range from 1.88 GeV to 209 GeV \cite{bethke}.
Below approximately 1 GeV, the coupling exceeds unity, and perturbative expansion in terms of the coupling no longer converges.
This poses a challenge for simulating hadronization, which takes place at this scale.

\section{Electroweak Theory}\label{sec:ewTheory}\label{sec:thyEW}
Electroweak theory (EW) is a field theory that describes the interactions between charged fermions and bosons.
Its predictions are invariant under $\text{SU}(2)\times\text{U}(1)$ group transformations.
The charges are \emph{weak isospin}, associated with the eigenvectors of the SU(2) generators, and \emph{weak hypercharge}, associated with U(1).
As is the case in QCD, the groups' generators are associated with bosons that help define covariant derivatives.
The three generators of SU(2), labeled $T^a$ for $a\in\{1,2,3\}$ are associated with three bosons $W_\mu^1$, $W_\mu^2$, and $W_\mu^3$.
The U(1) generator is associated with a boson $B_\mu$.
Each boson is a matrix valued 4-vector with a spacetime index $\mu$, and transforms $\text{SU}(2)\times\text{U}(1)$ group transformations as shown in Equation \ref{eqn:ewBosonTrans}.
The EW Lagrangian describes the interaction of these bosons with appropriately charged fermions.

For simplicity, the matrix field $\widetilde{W}_\mu\equiv\frac{\sigma_a}{2}W_\mu^a$ is defined where $\sigma_a$ are the Pauli matrices.
In this case, the transformations of the bosons under $\text{SU}(2)\times\text{U}(1)$ are given in Equation \ref{eqn:ewBosonTrans}.
\begin{equation}\begin{split}\label{eqn:ewBosonTrans}
    \text{SU}(2)\times\text{U}(1):\quad& B_\mu\to B_\mu+\frac{1}{g'}\partial_\mu\beta \\
    \text{SU}(2)\times\text{U}(1):\quad& \widetilde{W}^a_\mu\to U_L\widetilde{W}_\mu U^\dagger_L -\frac{i}{g}\partial_\mu U_LU^\dagger_L\\
\end{split}\end{equation}
Here the transformation matrix is $U_L=\exp{i\frac{\sigma_i}{2}\alpha^i(x)}$.
The four parameters specifying the gauge transformations are spacetime functions $\alpha^a(x)$ for $a\in\{1,2,3\}$ and $\beta(x)$.

The electroweak force acts on the chiral components of Dirac fermion fields.
These components, called left-handed and right-handed states, are defined from the projection of the parity operators:
\begin{equation}\begin{split}\label{eqn:chiralOps}
P_L=\frac{1-\gamma_5}{2} \quad\text{and}\quad
P_R=\frac{1+\gamma_5}{2}.
\end{split}\end{equation}
While these interactions are different for quarks and leptons, they share some similarity.
The fermion field can be decomposed into left- and right-handed component states which are denoted by $L$ and $R$ subscripts respectively.
The theory contains doublets under $\text{SU}(2)\times\text{U}(1)$ that consist of the left-handed projections of the fermions, given in Equation the left side of Table \ref{tab:su2Doublets}, and represented by $\psi_1$.
It also contains singlets that consist of the right-handed projections, represented by $\psi_2$ and $\psi_3$, which are given on the right side of the table.
\begin{table}[h!]
\begin{center}
\begin{tabular}{l l l l}
\toprule
Generation & Doublet & Singlets \\
\midrule
Quark 1 & $\psi_1=\begin{pmatrix}u\\d\end{pmatrix}_L$             & $\psi_2=u_R$, $\psi_3=d_R$ \\
Quark 2 & $\psi_1=\begin{pmatrix}s\\c\end{pmatrix}_L$             & $\psi_2=s_R$, $\psi_3=c_R$ \\
Quark 3 & $\psi_1=\begin{pmatrix}t\\b\end{pmatrix}_L$             & $\psi_2=t_R$, $\psi_3=b_R$ \\
Lepton 1 & $\psi_1=\begin{pmatrix}\nu_e\\e\end{pmatrix}_L$        & $\psi_2=\nu_{eR}$, $\psi_3=e_R$ \\
Lepton 2 & $\psi_1=\begin{pmatrix}\nu_\mu\\\mu\end{pmatrix}_L$    & $\psi_2=\nu_{\mu R}$, $\psi_3=\mu_R$ \\
Lepton 3 & $\psi_1=\begin{pmatrix}\nu_\tau\\\tau\end{pmatrix}_L$  & $\psi_2=\nu_{\tau R}$, $\psi_3=\tau_R$ \\
\bottomrule
\end{tabular}
\caption{Definitions of $\psi_1$ as left-handed doublets, and $\psi_2$ and $\psi_3$ as right-handed singlets, for different generations of fermion. To avoid confusion, it should be noted that there are no couplings to the right-handed neutrino fields in the Standard Model.}
\label{tab:su2Doublets}
\end{center}
\end{table}
The behavior of QCD may be discussed in terms of $\psi_1$, $\psi_2$, and $\psi_3$ before finally making reference to quarks, leptons or generations at the end.
It should be noted that since there is no compelling experimental evidence for the existence of right-handed neutrinos, and these will eventually have no impact on the final result.
These fields transform under group transformations as shown in Equation \ref{eqn:ewLepTrans}.
\begin{equation}\begin{split}\label{eqn:ewLepTrans}
    \text{SU}(2)_L:& \psi_1\to e^{-i\alpha^a\frac{\sigma^a}{2}}\psi_1 \\
    \text{SU}(2)_L:& \psi_{2,3}\to \psi_{2,3} \\
    \text{U}(1)_Y:& \psi_{1,2,3}\to e^{-iy\beta}\psi_{1,2,3} \\
\end{split}\end{equation}
\check
This introduces the conserved \emph{hypercharge} denoted $y$.
The subscript $L$ on the group indicates it acts only on right-handed fields.

As was done in the previous field theories, a covariant derivative acting on the lepton fields is defined:
\begin{equation}\begin{split}\label{eqn:ewCovDeriv}
        % D_\mu\begin{pmatrix}\nu_e\\e_L\end{pmatrix}=&[\partial_\mu-ig'B_\mu Y_{\ell_L}-igW^a_\mu T^a]\begin{pmatrix}\nu_e\\e_L\end{pmatrix} \\
        % D_\mu e_R=&[\partial_\mu-ig'B_\mu Y_{\ell_R}]e_R \\
        D_\mu\equiv& [\partial_\mu-ig'yB_\mu-igT^aW_\mu^a] ; \quad\text{Acting on }\psi_1\\
        D_\mu\equiv& [\partial_\mu-ig'yB_\mu]; \quad\text{Acting on }\psi_{2,3} \\
\end{split}\end{equation}
% Note! I found 2 refs with negative signs, one with positive signs. Since this is a coupling, I tend towards using negative...
The derivative acts differently on the left-handed doublets than on the right-handed singlets.
Here, $g'$ is the coupling associated with the U(1) subgroup, while $g$ is associated with the SU(2) subgroup.
Keep in mind there is still an implicit sum over the index $a$, and $T^a=\frac{\sigma^a}{2}$.
A field strength tensor for each of the four boson fields is defined and given in Equation \ref{eqn:ewFieldTensor}.
\begin{equation}\begin{split}\label{eqn:ewFieldTensor}
W^a_{\mu\nu} =& \partial_\mu W^a_\nu-\partial_\nu W^a_\mu-g\epsilon^{abc}W^b_\mu W^c_\nu; \quad a\in\{1,2,3\} \\
B_{\mu\nu} =& \partial_\mu B_\nu-\partial_\nu B_\mu \\
\end{split}\end{equation}
Here, $\epsilon^{abc}$ is the SU(2) structure constant.
An analogous term vanishes in the definition of $B^a_{\mu\nu}$ because U(1) is Abelian with structure constants of zero.
Again for simplicity, the definition $\widetilde{W}_{\mu\nu}\equiv\frac{\sigma_a}{2}W_{\mu\nu}^a$ is made.


% Lagrangian
The field strength tensors combine to contract to produce the product $-\frac{1}{4}W^{a,\mu\nu}W^{a}_{\mu\nu}-\frac{1}{4}B^{\mu\nu}B_{\mu\nu}$.
The covariant derivative in Equation \ref{eqn:ewCovDeriv} produces terms $i\overline{\psi}_j\slashed{D}\psi_j$ for each field $\psi$ in the theory.
Combined, these terms form the EW Lagrangian given in Equation \ref{eqn:ewNoInt}.
\begin{equation}\begin{split}\label{eqn:ewNoInt}
\mathcal{L} = -\frac{1}{4}W^{a,\mu\nu}W^{a}_{\mu\nu}- \frac{1}{4}B^{\mu\nu}B_{\mu\nu}+ \sum_j^3 i\overline{\psi}_j\slashed{D}\psi_j
\end{split}\end{equation}
Here, the form of the covariant derivative reveals two things.
First, noting the covariant derivative's dependence in Equation \ref{eqn:ewCovDeriv}, there are no terms containing right-handed fermions and either $W^1_\mu$ or $W^2_\mu$. Since these form the basis for the \W bosons, this produces no interaction with the right-handed component of fermion fields.
Second, while left-handed fermions appear in terms with both $B_\mu$ and $W^3_\mu$, right-handed fermions appear only in terms with $B_\mu$. Therefore the \Z boson interacts differently with the left- and right-handed fields.
Together, these facts show that EW theory is a \emph{chiral} theory in that it treats the left- and right-handed components of fields differently.

The fermion mass term $m\overline{\psi}\psi$ is conspicuously missing from Equation \ref{eqn:ewNoInt}.
This is because the left (doublet) and right (singlet) components of the fermion fields transform differently under SU(2) gauge transformations, as shown in Equation \ref{eqn:ewLepTrans}, therefore the product transforms as a doublet.
The EW Lagrangian cannot include such terms and remain a singlet, and so these are forbidden.
The fermions will instead gain their mass through the Brout-Englert-Higgs mechanism described in Section \ref{sec:higgsMechanism}.

Linear combinations of $B_\mu$, $W_\mu^1$, $W_\mu^2$, and $W_\mu^3$ produce the observed $A_\mu$, \Z, and \W bosons.
The mixing of $B_\mu$ and $W_\mu^3$ described by the Weinberg angle $\cos\theta_W=M_Z/M_W$, and is given in Equation \ref{eqn:wbmix}.
\begin{equation}\begin{split}\label{eqn:wbmix}
    % \begin{pmatrix}Z_\mu\\A_\mu\end{pmatrix}=&\begin{pmatrix}\cos{\theta_W}&-\sin{\theta_W}\\\sin{\theta_W}&\cos{\theta_W}\end{pmatrix} \begin{pmatrix}W^3_\mu\\B_\mu\end{pmatrix}
    Z_\mu=\cos\theta_W W^3_\mu-\sin\theta_W B_\mu \\
    A_\mu=\sin\theta_W W^3_\mu+\cos\theta_W B_\mu \\
\end{split}\end{equation}
The Weinberg angle is measured to be $\sin^2\theta_W$=0.231 at \Z mass scale.
Meanwhile the fields $W_\mu^1$ and $W_\mu^2$ combine to yield the charged \W bosons as shown in Equation \ref{eqn:wwmix}.
\begin{equation}\begin{split}\label{eqn:wwmix}
    \W_\mu=\frac{1}{\sqrt{2}}(W_\mu^1\mp iW_\mu^2)
\end{split}\end{equation}
The EW Lagrangian can eventually be re-written in terms of these mass eigenstates.
Since Feynman diagrams are usually drawn in that context, the Feynman rules will be defined after this.
There are a few useful points to be made meanwhile about the coupling strengths.
Plug Equations \label{eqn:wbmix} solution for $B_\mu$ into the right-handed covariant derivative in Equation \ref{eqn:ewCovDeriv}'s reveals that the right-handed lepton fields have a coupling to the $A_\mu$ field of $-g'\cos\theta_W$.
Meanwhile a comparison with electrodynamics identifies this as the electric coupling strength.
Requiring these to be equal necessitates $g=e/\cos\theta_W=0.357$.
An analogous requirement for electrically neutral neutrinos yields $g=g'/\tan\theta_W=0.652$.

Before continuing, an electromagnetic charge operator $Q_i$ is defined for each $\psi_i$.
$Q_2$ and $Q_3$ are simply the electromagnetic charges of $\psi_2$ and $\psi_3$.
$Q_1$ is a diagonal matrix $\texttt{diag}\{a,b\}$ where $a$ and $b$ are the electromagnetic charges of the upper and lower components of the $\psi_1$ doublet.
This allows the definition of hypercharge $y_i$ each state $\psi_i$: $Y\equiv Q-T_3$.
$T_3$ is the weak isospin of the left-handed fermions, corresponding to their position in the doublet. It is zero for the right-handed fermions.
\cite{pich}

The Lagrangian in Equation \ref{eqn:ewNoInt} contains interaction terms between fermions and the charged and neutral bosons.
These can be seen by expanding the covariant derivative and plugging in the inverse of Equations \ref{eqn:wbmix} and \ref{eqn:wwmix}.
The charged current takes the form in Equation \ref{eqn:ewLagrangianCc}.
\begin{equation}\begin{split}\label{eqn:ewLagrangianCc}
    \mathcal{L}_\text{CC} = -\frac{g}{\sqrt{2}}W_\mu^+[\nu_\ell\gamma^\mu \ell_L+\ubar\gamma^\mu d_L]
\end{split}\end{equation}
Here there is an implicit sum over lepton and quark generations, with $\ubar$ and $d_L$ corresponding to up and down type quarks from each generation.

The Lagrangian in Equation \ref{eqn:ewNoInt} also contains interactions between fermions and the neutral bosons $A_\mu$ and $Z_\mu$.
These are similarly identified by plugging equations Equations \ref{eqn:wbmix} and \ref{eqn:wwmix} (inverted) into the covariant derivative in Equation \ref{eqn:ewCovDeriv}.
Grouping the terms with $A_\mu$ and $Z_\mu$ yields the Lagrangian of neutral current in Equation \ref{eqn:ewLagrangianNc}.
\begin{equation}\begin{split}\label{eqn:ewLagrangianNc}
    \mathcal{L}_\text{NC} = \sum_j^3 \overline{\psi}_j\gamma^\mu\left[A_\mu(gT_3\sin\theta_W+g'y_j\cos\theta_W)+Z_\mu(gT_3\cos\theta_W-g'y_j\sin\theta_W)\right]
\end{split}\end{equation}
The neutral current Lagrangian can be broken down further to reveal the QED interactions.
The constraints on the Weinberg angle, and the definition $Y\equiv Q-T_3$ allow the term containing $A_\mu$ to produce a Lagrangian describing fermion-photon interactions:
\begin{equation}\begin{split}
    \mathcal{L}_{Af\fbar}=-eA_\mu\sum_j\overline{\psi}_j\gamma^\mu Q_j\psi_j.
\end{split}\end{equation}
The latter part of Equation \ref{eqn:ewLagrangianNc} describes fermion-\Z boson interactions, and is written:
\begin{equation}\begin{split}
    \mathcal{L}_{Zf\fbar} = -\frac{g}{2\cos\theta_W}Z_\mu\sum_f\overline{\psi}_j\gamma^\mu(\sigma_3-2Q_j\sin^2\theta_W)\psi_j.
\end{split}\end{equation}

In summary, the QCD Lagrangian contains terms describing three-point interactions between a each vector boson and two fermions.
Further interactions appear in the following section.

% \begin{table}[H]
% % \begin{table}[htp]
% \begin{center}
% {\footnotesize
% \begin{tabular}{l | l | l l l}
% \toprule
% Name & Lagrangian Term & Drawing & Rule \\
% \midrule
% \centered{\W-lepton\\interaction} & $-\frac{g}{\sqrt{2}}W_\mu^+\nu_\ell\gamma^\mu\ell_L$ &  $\feynmandiagram [small,baseline=(d.base), horizontal=d to b] { a[particle=\(\ell/\nu_\ell\)] -- [fermion] b -- [fermion] c[particle=\(\nu_\ell/\ell\)], b -- [boson] d [particle=\(\W\)], };$ & $-i\frac{g}{\sqrt{2}}P_L\gamma^\mu$  \\[1.0em]
% \centered{\W-quark\\interaction} & \centered{$-\frac{g}{\sqrt{2}}W_\mu^+\ubar\gamma^\mu d_L$\\($u/d$ is an up/down-type quark)} &  $\feynmandiagram [small,baseline=(d.base), horizontal=d to b] { a[particle=\(d_\beta/u_\alpha\)] -- [fermion] b -- [fermion] c[particle=\(u_\alpha/d_\beta\)], b -- [boson] d [particle=\(\W\)], };$ & $-i\frac{g}{\sqrt{2}}P_L\gamma^\mu V_{\alpha\beta}$  \\[1.0em]
% \centered{\Z-fermion\\interaction} & $-\frac{g}{2\cos\theta_W}Z_\mu\overline{\psi}_j\gamma^\mu(v_f-a_f\gamma^5)\psi_j$ &  $\feynmandiagram [small,baseline=(d.base), horizontal=d to b] { a -- [fermion] b -- [fermion] c, b -- [boson] d [particle=\(Z\)], };$ & $-i\frac{g}{2\cos\theta_W}(X^f)\gamma^\mu$  \\[1.0em]
% \centered{Photon-fermion\\interaction} & $-eA_\mu\overline{\psi}_j\gamma^\mu Q_j\psi_j$ &  $\feynmandiagram [small,baseline=(d.base), horizontal=d to b] { a -- [fermion] b -- [fermion] c, b -- [boson] d [particle=\(\gamma\)], };$ & $-ieQ_j\gamma^\mu$  \\[1.0em]
% % gff \\
% \bottomrule
% \end{tabular}
% }
% \caption{Summary of the Feynman rules for the charged and neutral EW interactions. Note the $P_L$ appears in charged current interactions, but not the neutral current interactions. Several abbreviations are used: $a_f=T_3^f$, $v_f=T_3^f(1-4|Q_f|\sin^2\theta_W)$, and $X^f=T_3^f(1-4|Q_f|\sin^2\theta_W-\gamma^5)$. The matrix $V_{\alpha\beta}=\ident$, but the fields and $V_{\alpha\beta}$ will be replaced in the following sections.}
% \label{tab:ewRules0}
% \end{center}
% \end{table}

\subsection{Electroweak Symmetry Breaking}\label{sec:higgsMechanism}

In the previous three subsections, the QED, QCD, and EW Lagrangians are invariant under U(1), SU(3), and $\text{SU}(2)\times\text{U}(1)$ group transformations respectively.
All are invariant under \poincare group transformations as well.
The physical state of a system, however, is not necessarily invariant under these transformations.
A classic example of this is found in a cooling magnet.
Above the Curie temperature, the component atoms' magnetite dipole moments are oriented isotropically, and hence the system is invariant under rotations.
When the material cools, the spins align in a particular direction, thereby breaking this invariance.
The physical laws remain invariant, but the state of the system has removed rotational invariance.
This situation is referred to as \emph{symmetry breaking}.

A new doublet under SU(2) is introduced, with complex scalar components $\phi^+$ and $\phi^0$: $\Phi=\begin{pmatrix}\phi^+&\phi^0\end{pmatrix}^T$.
This field transforms under the EW guage transformations as:
\begin{equation}\begin{split}\label{eqn:scalarTransform}
    \text{SU}(2)_L\quad\Phi(x)\to\Phi'(x)=&e^{-i\theta^a(x)\sigma^a/2}\Phi(x) \\
    \text{U}(1)\quad\Phi(x)\to\Phi'(x)=&e^{-i\theta^a(x)/2}\Phi(x). \\
\end{split}\end{equation}
Along with $\Phi$, a potential energy is added to the system as a function of the field:
\begin{equation}\begin{split}
V(\Phi,\Phi^\dagger)=m^2\Phi^\dagger\Phi+\lambda(\Phi^\dagger\Phi)^2,
\end{split}\end{equation}
where $m^2$ and $\lambda$ are undetermined constants.
If $m^2<0$ and $\lambda>0$ then the potential $V(\Phi,\Phi^\dagger)$ is minimized when $\Phi^\dagger\Phi=-\half m^2/\lambda$.
This defines the vacuum expectation value, or VEV, of the field $\Phi$ as $v=\sqrt{-m^2/\lambda}$.
By convention, the VEV is allocated to the neutral field $\phi^0$ by a choice of an SU(2) gauge transformation: $\braket{0|\Phi|0}=\begin{pmatrix}0&v/\sqrt{2}\end{pmatrix}^T$.

The four degrees of freedom of $\Phi$ can be described by real scalar fields $h$ and $G^a$ with $a\in\{1,2,3\}$, as shown in the first line of Equation \ref{eqn:scalarExpand}.
The field $h$ is called the Higgs field.
The fields $G^a$ are pseudo Nambu-Goldstone bosons, and can be removed by a gauge choice $\theta^a=-G^a(x)/v$ from Equation \ref{eqn:scalarTransform}.
This is shown in the second line of Equation \ref{eqn:scalarExpand}.
\begin{equation}\begin{split}\label{eqn:scalarExpand}
\Phi(x)=&e^{iG^a(x)\sigma^a/2v}\begin{pmatrix}0\\\frac{v+h(x)}{\sqrt{2}}\end{pmatrix} \\
\Phi(x)=&\begin{pmatrix}0\\\frac{v+h(x)}{\sqrt{2}}\end{pmatrix}; \quad\text{After gauge choice} \\
\end{split}\end{equation}

The new field $\Phi$ allows new terms in the EW Lagrangian.
First, a covariant derivative is defined:
\begin{equation}\begin{split}
        D_\mu\Phi=&\frac{1}{\sqrt{2}}\begin{pmatrix}0\\\partial_\mu h\end{pmatrix}+\frac{i}{\sqrt{2}}\left[\frac{g'}{2}B_\mu+\frac{g}{2}W_\mu^a\sigma^a\right] \begin{pmatrix}0\\v+h\end{pmatrix}
\end{split}\end{equation}
This defines an invariant kinetic term for $\Phi$, which is expanded in the first line of Equation \ref{eqn:scalarKinetic}.
\begin{equation}\begin{split}\label{eqn:scalarKinetic}
    D^\mu\Phi^\dagger D_\mu\Phi =& \half\partial_\mu h\partial^\mu h+\frac{(v+h)^2}{2}\begin{pmatrix}0&1\end{pmatrix}\left[\frac{g'}{2}B+\frac{g}{2}W^a\sigma\right]^2\begin{pmatrix}0\\1\end{pmatrix} \\
    =& \half\partial_\mu h\partial^\mu h+\frac{(v+h)^2}{4}\left[g^sW^+_\mu W^{-\mu}+\half(g^2+g'^2)Z_\mu Z^\mu\right] \\
\end{split}\end{equation}
In the second line, the fields $B$ and $W_\mu^a$ are replaced with the vector fields from Equations \ref{eqn:wbmix} and \ref{eqn:wwmix}.
In this form, three things become apparent after expanding $(v+h)^2$ and looking at terms, including $h$.
First, terms involving two vector fields and one Higgs field appear, leading to three-point vertices.
Second, terms involving two vector fields and two Higgs field appear, leading to four-point vertices.
Third, no term appears linking the photon field $A_\mu$ directly to the Higgs field.

Next, expanding $(v+h)^2$ includes terms proportional to $v^2$, including only $W^+_\mu W^{-\mu}$ or $Z_\mu Z^\mu$ fields.
The Higgs mechanism has provided invariant mass terms for the \Z and \W bosons.
The masses are easily ready from the coefficients:
\begin{equation}\begin{split}
m_{\W}^2=&\frac{v^2}{4}g^2 \\
m_{\Z}^2=&\frac{v^2}{4}(g^2+g'^2) \\
\end{split}\end{equation}
No mass term appears for the photon field.

The Higgs field also appears in the potential energy $V(\Phi,\Phi^\dagger)$.
Noting $m^2=-v^2\lambda$ and inserting the form for $\Phi$ from Equation \ref{eqn:scalarExpand} into the equation for the potential yields:
\begin{equation}\begin{split}
    V(\Phi,\Phi^\dagger)=\lambda v^2h^2+\lambda vh^3+\frac{\lambda}{4}h^4.
\end{split}\end{equation}
This identifies three and four point Higgs self interaction vertices.
Additionally, the potential includes a mass term for the Higgs boson: $m_h=\sqrt{2\lambda}v$.

The final result is the origin of leptonic mass.
The mass term that appears in the QED Lagrangian ($m\overline{\Psi}\Psi$) is not an SU(2) singlet.
Instead, consider the Yukawa coupling between left ($L_i=\begin{pmatrix}\overline{\nu}_\ell\overline{\ell}_L\end{pmatrix}$) and right-handed ($\ell_R$) leptons $i$ with the scalar field $\Phi$.
\begin{equation}\begin{split}\label{eqn:leptonYukawaLagrangian}
        \mathcal{L}_{\text{Yukawa}}=&-y_i L_i\Phi R_i \\
        =&-y_i\begin{pmatrix}\overline{\nu}_\ell&\overline{\ell}_L\end{pmatrix}\begin{pmatrix}\phi^+\\\phi^0\end{pmatrix}\ell_R \\
        % =&-y_i\begin{pmatrix}\overline{\nu}_e&\overline{e}_L\end{pmatrix}\Phi e_R \\
        =&-\frac{y_iv}{\sqrt{2}}\overline{\ell}\ell-\frac{y_i}{\sqrt{2}}\overline{\ell}\ell h \\
\end{split}\end{equation}
The step from the second line to third line corresponds to the gauge choice in Equation \ref{eqn:scalarExpand}.
This term is invariant under SU(2) transformations, because the exponential term in Equation \ref{eqn:scalarTransform} is canceled by the analogous term for the doublet in Equation \ref{eqn:ewLepTrans}.
As a result of the Yukawa coupling, a mass term has appeared for the lepton $i$:
\begin{equation}\begin{split}\label{eqn:yukawaMass}
    m_\ell=\frac{y_iv}{\sqrt{2}}.
\end{split}\end{equation}
In addition, a three point interaction between the lepton and Higgs appears, with a coupling:
\begin{equation}\begin{split}\label{eqn:yukawaCoupling}
    g_{h\ell\ell}=\frac{y_i}{\sqrt{2}}.
\end{split}\end{equation}
The other fermions, excluding neutrinos, gain their mass through similar Yukawa couplings\footnote{Quarks benefit from additional mass from chiral symmetry breaking.}.
Equations \ref{eqn:yukawaMass} and \ref{eqn:yukawaCoupling} are particularly relevant to the subject of this thesis.
The former shows that the muon gains its mass due to the scalar field $\Psi$, and the latter is the topic of measurement in Chapter \ref{sec:hmumu}.

% The addition of the scalar field to the Lagrangian adds new Feynman rules, and these are summarized in Tables \ref{tab:ewRules1} and \ref{tab:ewRules2}.


% \afterpage{
% \begin{table}[H]
% % \begin{table}[htp]
% \begin{center}
% {\footnotesize
% \begin{tabular}{l | l | l l l}
% \toprule
% Name & Lagrangian Term & Drawing & Rule \\
% \midrule
% \centered{\W propagator}      &  & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=a to b] { a[particle=\(\mu\)] --[charged boson] b[particle=\(\nu\)] }; $} & \centered{$\frac{i}{p^2-m_W^2+i\epsilon}\left[-g_{\mu\nu}+\frac{p_\mu p_\nu}{m_W^2}\right]$} \\ [1.0em]
% \centered{\Z propagator}      &  & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=a to b] { a[particle=\(\mu\)] --[charged boson] b[particle=\(\nu\)] }; $} & \centered{$\frac{i}{p^2-m_Z^2+i\epsilon}\left[-g_{\mu\nu}+\frac{p_\mu p_\nu}{m_Z^2}\right]$} \\ [1.0em]
% \centered{Fermion-Higgs\\interaction}      & $\frac{y_i}{\sqrt{2}}\overline{\ell}\ell h$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=d to v] { {a[particle=\(\ell\)]} --[fermion] v , v --[fermion]  b[particle=\(\ell\)] , {d[particle=\(h\)]} --[scalar] v  };$} & \centered{$-iy_\ell/\sqrt{2}$} \\ [1.0em]
% \centered{$W^+W^-H$\\interaction}      & $\frac{v}{2}g_sW^+_\mu W^{-\mu}$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=d to v] { {a[particle=\(W^\nu\)]} --[charged boson] v , v --[charged boson]  b[particle=\(W^\mu\)] , {d[particle=\(h\)]} --[scalar] v  }; $} & \centered{$ig^{\mu\nu}gv/2$} \\ [1.0em]
% \centered{$ZZH$\\interaction}      &  $\frac{v}{4}(g^2+g'^2)Z_\mu Z^\mu$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=d to v] { {a[particle=\(Z^\nu\)]} --[charged boson] v , v --[charged boson]  b[particle=\(Z^\mu\)] , {d[particle=\(h\)]} --[scalar] v  }; $} & \centered{$ig^{\mu\nu}(g^2+g'^2v/2$)} \\ [1.0em]
% \centered{Three Higgs\\interaction}        & $\lambda vh^3$  & \centered{$\feynmandiagram [ssmall,baseline=(d.base), horizontal=d to v] { c[particle=\(h\)] -- [scalar] v , a[particle=\(h\)] -- [scalar] v, d[particle=\(h\)] -- [scalar] v , };$} & \centered{$-i6\lambda v$} \\ [1.0em]
% \centered{Four Higgs\\interaction}         &  $\frac{\lambda}{4}h^4$ & \centered{$\feynmandiagram [ssmall,baseline=(d.base), horizontal=a to c] { {a[particle=\(h\)],b[particle=\(h\)]} --[scalar] v --[scalar] {c[particle=\(h\)],d[particle=\(h\)]} };$} & \centered{$-i6\lambda$}  \\[1.0em]
% \centered{$WWHH$\\interaction}             & $\frac{h^2}{4}g_sW^+_\mu W^{-\mu}$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=a to c] { {a[particle=\(W^\mu\)]} --[charged boson] v --[charged boson] b[particle=\(W^\nu\)], v --[scalar] {c[particle=\(h\)],d[particle=\(h\)]} };$} & \centered{$ig^{\mu\nu}g^2/2$}  \\[1.0em]
% \centered{$ZZHH$\\interaction}             & $\frac{h^2}{8}(g^2+g'^2)Z_\mu Z^\mu$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=a to c] { {a[particle=\(Z^\mu\)]} --[charged boson] v --[charged boson] b[particle=\(Z^\nu\)], v --[scalar] {c[particle=\(h\)],d[particle=\(h\)]} };$} & \centered{$ig^{\mu\nu}(g^2+g'^2)/2$}  \\[1.0em]
% \bottomrule
% \end{tabular}
% }
% \caption{Summary of the Feynman rules for the EW Lagrangian, including the scalar Higgs field. }
% \label{tab:ewRules1}
% \end{center}
% \end{table}
% \clearpage
% }


% \afterpage{
% \begin{table}[H]
% % \begin{table}[htp]
% \begin{center}
% {\footnotesize
% \begin{tabular}{l | l | l l l}
% \toprule
% Name & Lagrangian Term & Drawing & Rule \\
% \midrule
% \centered{Charged-current quark\\interaction}  & \centered{$\frac{g}{\sqrt{2}}W_\mu^+[\ubar\gamma^\mu d_L]$\\($u/d$ is an \\up/down-type quark)}     & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=d to v] { {a[particle=\(W^\mu\)]} --[charged boson] v , v --[fermion] b[particle=\(q\)], {d[particle=\(q'\)]} --[fermion] v  };$} & $\frac{g}{2^{3/2}}(1-\gamma_5)$  \\[1.0em]
% \centered{Charged-current lepton\\interaction} & $\frac{g}{\sqrt{2}}W_\mu^+\nu_\ell\gamma^\mu \ell_L$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=d to v] { {a[particle=\(W^\mu\)]} --[charged boson] v , v --[fermion] b[particle=\(\ell\)], {d[particle=\(\nu_\ell\)]} --[fermion] v  };$} & $\frac{g}{2^{3/2}}(1-\gamma_5)$  \\[1.0em]
% \centered{$\gamma W^+W^-$\\interaction}        & $\mathcal{L}_3$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=d to v] { {a[particle=\(W^\mu\)]} --[charged boson,momentum=\(k\)] v , v --[charged boson,momentum=\(q\)] b[particle=\(W^\nu\)], {d[particle=\(A_\mu\)]} --[boson,momentum=\(p\)] v  };$} & \centered{$ie[g^{\mu\nu}(p-q)^\rho+$\\$~~~g^{\nu\rho}(q-k)^\mu+$\\$~~~g^{\rho\mu}(k-p)^\nu]$}  \\[1.0em]
% \centered{$Z W^+W^-$\\interaction}             & $\mathcal{L}_3$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=d to v] { {a[particle=\(W^\mu\)]} --[charged boson,momentum=\(k\)] v , v --[charged boson,momentum=\(q\)] b[particle=\(W^\nu\)], {d[particle=\(Z_\mu\)]} --[boson,momentum=\(p\)] v  };$} & \centered{$ig\cos(\theta_W)[g^{\mu\nu}(p-q)^\rho+$\\$~~~~~~~~~~~~~~g^{\nu\rho}(q-k)^\mu+$\\$~~~~~~~~~~~~~~g^{\rho\mu}(k-p)^\nu]$}  \\[1.0em]
% \centered{$W^+W^-W^+W^-$\\interaction}         & $\mathcal{L}_4$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=a to c] {{a[particle=\(W^\sigma\)]} --[charged boson] v --[charged boson] b[particle=\(W^\mu\)], c[particle=\(W^\rho\)] --[charged boson] v --[charged boson] d[particle=\(W^\nu\)] }; $} & \centered{$ig^2X^{\mu\nu,\rho\sigma}$}  \\[1.0em]
% \centered{$\gamma\gamma WW$\\interaction}      & $\mathcal{L}_4$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=a to c] {{a[particle=\(A^\nu\)]} --[boson] v --[boson] b[particle=\(A^\mu\)], c[particle=\(W^\rho\)] --[charged boson] v --[charged boson] d[particle=\(W^\sigma\)] }; $} & \centered{$-ie^2X^{\mu\nu,\rho\sigma}$}  \\[1.0em]
% \centered{$ZZW^+W^-$\\interaction}             & $\mathcal{L}_4$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=a to c] {{a[particle=\(Z^\nu\)]} --[charged boson] v --[charged boson] b[particle=\(Z^\mu\)], c[particle=\(W^\rho\)] --[charged boson] v --[charged boson] d[particle=\(W^\sigma\)] }; $} & \centered{$-ig^2\cos^2(\theta_W)X^{\mu\nu,\rho\sigma}$}  \\[1.0em]
% \centered{$\gamma ZWW$\\interaction}           & $\mathcal{L}_4$ & \centered{$\feynmandiagram [ssmall,baseline=(v.base),horizontal=a to c] {{a[particle=\(Z^\nu\)]} --[charged boson] v --[charged boson] b[particle=\(A^\mu\)], c[particle=\(W^\rho\)] --[charged boson] v --[charged boson] d[particle=\(W^\sigma\)] }; $} & \centered{$-ig^2\cos(\theta_W)\sin(\theta_W)X^{\mu\nu,\rho\sigma}$}  \\[1.0em]
% \bottomrule
% \end{tabular}
% }
% \caption{Summary of the Feynman rules for the EW Lagrangian, including the scalar Higgs field. For convenience, $X^{\mu\nu,\rho\sigma}\equiv2g^{\mu\nu}g^{\rho\sigma}-g^{\mu\rho}g^{\nu\sigma}-g^{\mu\sigma}g^{\nu\rho}$. Fields marked ``$\mathcal{L}_3$'' and ``$\mathcal{L}_4$'' arise from the expansion of Equation \ref{eqn:ewNoInt}.}
% \label{tab:ewRules2}
% \end{center}
% \end{table}
% \clearpage
% }

\subsection{Cabibbo-Kobayashi-Maskawa mixing}
% \subsection{Chiral Symmetry Breaking}

The Yukawa Lagrangian in Equation \ref{eqn:leptonYukawaLagrangian} can be expanded in a compact form for the nine charged fermions, shown in Equation \ref{eqn:fermionYukawaLagrangian}
\begin{equation}\begin{split}\label{eqn:fermionYukawaLagrangian}
        \mathcal{L}_{\text{Yukawa}}=-\left(1+\frac{h}{v}\right)\left[ \overline{\ell'}_L^i[m_{\ell}]_i^j\ell'_{Rj} + \overline{u'}_L^i[m_{u}]_i^ju'_{Rj} + \overline{d'}_L^i[m_{d}]_i^jd'_{Rj} \right]
\end{split}\end{equation}
Here, the leptons, up-type quarks, down-type quarks are grouped into vectors $\ell'$, $u'$, and $d'$ containing fields for each generation.
These are the \emph{mass-eigenstate} fields $f$, and are related to the primed fields $f'$ by multiplication by unitary matrices $U_L$ and $U_R$: $f_L'=U_Lf_L$ and $f'_R=U_Rf_R$.
The $3\times3$ matrices $[m_{\ell}]$, $[m_{u}]$, and $[m_{d}]$ by analogy to Equation \ref{eqn:leptonYukawaLagrangian} represent the masses fermions.
The unitary matrices in $f'_L$ and $f'_R$ are chosen, such that the mass matrices are diagonal through the multiplication $U^\dagger_L[m]U_R$.
The result is that Equation \ref{eqn:fermionYukawaLagrangian} is simplified such that mass terms appear as in Equation \ref{eqn:fermionYukawaLagrangian}.

The electroweak Lagrangian in Equation of the form \ref{eqn:ewNoInt} is nearly unchanged with the substitution of the mass-eigenstate fields.
The exception is the interaction term between \W bosons and quarks, as given in Equation \label{eqn:ewLagrangianCc}.
The change to mass-eigenstates introduces the product between two unrelated unitary matrices $U$ and $U'$ that correspond to the up-type and down-type quarks.
The product $U^\dagger U'$ is not one, but instead defines the unitary Cabibbo-Kobayashi-Maskawa (CKM) matrix $V$.
Instead of the choice of a unitary matrix $f_L'=U_Lf_L$ and $f'_R=U_Rf_R$ that is used to define the mass-eigenstates for leptons, an analogous definition for quarks is made: $q'_{L,i}=V_i^jq_{L,j}$.
The Lagrangian may be rewritten in terms of the mass-eigenstates with the use of the CKM matrix:

\begin{equation}\begin{split}
    \mathcal{L}=-\frac{g}{\sqrt{2}}V_i^jW_\mu\ubar^i_L\gamma d_{L,j}.
\end{split}\end{equation}

The global measurement of the CKM matrix parameters, along with unitary within generations, is given in Equation \ref{eqn:ckm}. \cite{pdg2019}
\begin{equation}\begin{split}\label{eqn:ckm}
V=
\begin{pmatrix}
0.97446 \pm 0.00010 & 0.22452 \pm 0.00044 & 0.00365 \pm 0.00012 \\
0.22438 \pm 0.00044 & 0.97359^{+0.00010}_{-0.00011} & 0.04214 \pm 0.00076 \\
0.00896^{+0.00024}_{-0.00023} & 0.04133 \pm 0.00074 & 0.999105 \pm 0.000032
\end{pmatrix}
\end{split}\end{equation}


\section{Summary}

Electroweak symmetry breaking is the final piece of the Standard Model.
In this section, the full SM Lagrangian is constructed, and the remaining Feynman rules are presented.

Since the discovery of the Higgs boson, the full SM Lagrangian is often presented in the compact form of Equation \ref{eqn:smLagrangian}.
\begin{flalign}\label{eqn:smLagrangian}
~&~& \mathcal{L}_\text{SM} =& -\frac{1}{4}F^a_{\mu\nu}F^{a,\mu\nu}               &\text{gauge fields} &~ \notag\\
~&~&                        & +i\overline{\Psi}_i\slashed{D}\Psi_i+\text{h.c.}   &\text{covariant derivatives} &~ \notag\\
~&~&                        & +\overline{\Psi}_iy_{ij}\Psi_j\Phi+\text{h.c.}                &\text{Yukawa couplings} &~ \notag\\
~&~&                        & +D_\mu\Phi D^\mu\Phi-V(\Phi,\Phi^\dagger);         &\text{Higgs terms} &~
\end{flalign}
This form of the Lagrangian begins with a sum over the contracted the gauge field strength tensors $F^a$, which includes the eight gluon fields from Equation \ref{eqn:qcdLagrangian2} and the four electroweak fields from Equation \ref{eqn:ewFieldTensor}.
The next line contains the covariant derivatives acting on fermion fields $\Psi_i$. There is an implicit sum over index $i$ covering all fermions.
The third line contains Yukawa couplings for each fermion, again summed over indices $i$ and $j$.
The Yukawa coupling constants are described compactly by $3\times3$ matrices, $y_{ij}=\frac{\sqrt{2}}{mv}$, that correspond to the masses from Equation \ref{eqn:fermionYukawaLagrangian}.
This is because the fields $\Psi$ are gauge eigenstates, rather than mass eigenstates.
The final line contains the scalar kinetic term and the Higgs potential.

The previous section dealt with the abstract notions of groups and relations.
Representations of these groups provide a useful description of the physical world.
This section will discuss how these representations are used in particle physics theory.
The task is to build a map between the mathematical structures, which are stripped of physical meaning, and the physical model.

The connection between the abstract words of groups and symmetries and the more concrete world of QFT is as follows.
The fields of fermions correspond to the modules of certain representations, on which the members of the algebra can act.
Charges, the physical characteristics of particles that are conserved, correspond to the eigenvalues of the group's generators.
The number of charges described by a group is equal to the dimension of the representation.
Physical interactions through a force are described by the eigenvectors of the earlier defined Cartan generators of the group.
The generators of the group describe particles that mediate forces to particles charged under the group.
The force-carrying particles described by the Cartan generators do not change the corresponding charge of a particle.
Non-Cartan generators, however, can change charge.

