\begin{center}
\textbf{\LARGE Acknowledgements}
\end{center} 

During my time working on this project, I have benefitted from the help and generosity of many people.
I would like to thank the members the ATLAS Collaboration who have built and run the detector and computing infrastructure that has enabled my work.
I would also like to thank the enginners and staff at CERN who help operate the LHC and maintian the site.
I am thankful to the Michigan professors Junjie Zhu, Jianming Qian, Homer Neal, and Jay Chapman for building a supportive university environment.
I am thankful to the many friends I have made with for engaging in interesting discussions, especially Tom Cheng, Yicheng Guo, Zhi Zheng, Zirui Wang, Rongkun Wang, Zhongyukun Xu, and Siyuan Sun, Yanlin Liu, Hao Liu, and Dan Marley.
In particular, Matthew Klein has provided patient and invaluable supervision on topics of data analysis.
During my authorship qualification, I benifitted from the guidance of Tom Schwarz and technical instruction of Xueye Hu, Jinhong Wang, and Bob Ball.
On the search for the Higgs decay to two muons, I am thankful for the guidance of guidance of Yusheng Wu, Yanlin Liu, and Bing Li.
On the search for contact interactions, I would like to thank the Exotics-LPX Dilepton group and our ever patient editorial board, Paul de Jong, Ulla Blumenschein, and Elliot Lipelles for their feedback.
I would especially like to thank Uta Klein, Dan Hayden, Etienne Dreyer, Peter Falke, Tanya Hryn'Ova, Sean Lawler for their contributions and guidance.
I am thankful for the LPX and Exotics group leadership of Yasu Okumura, Nikolina Llic, Antonio Sidoti, Carl Gwilliam, and Oliver Stelzer-Chilton.
Penultimately, I want to note my great fortune in working on this project with Deshan Abhayasinghe and Noam Tal Hod, who have made the this project one of the highlights of the past five years..
Finally, I want to my deepest thanks to my advisor, Bing Zhou.
She has worked tirelessly to support me since, as an undergraduate, I enrolled in her section of Physics 405: Intermediate Electromagnitism.
As a role model, her presistant curiosity and attention to detail have provided an exceptional example of an experimental physicist. 
As a mentor, she has always pushed me to be better regardless of my mistakes. 


